An Audio Interface for an Active Perception Navigation Aid for Blind People

### 2019-05-13 article.tex
Removed characterisation experiments

\begin{abstract}
	Our aim is to build a portable active navigation system for the visually impaired that uses active perception techniques and a combination of feedback modes to identify a scene and guide the user to his/her destination.
	In this paper, we investigate the effectiveness of a spatial audio tone with a varying pitch component, played through bone-conducting headphones, in conveying the pan and tilt angles of a target to the user in a pointing task.
	We also wish to see how changes to the rate of change of the pitch affects the user's performance.
	We conducted a set of experiments with blindfolded and visually impaired users and found that the varying pitch component gives an average error of approximately \SI{20}{\degree} when searching for a target; in line with previous results from other authors.
	Furthermore, we were able to demonstrate that the audio interface adheres to Fitts's Law and used it as a metric to determine which pitch setting produces the best results.
	We discovered a trade-off between the speed and accuracy in the pointing task, which are maximised when the tone's pitch is adjusted to a low and high rate of change respectively. 
\end{abstract}

@@

In recent years, governments have spearheaded numerous initiatives to support people with disabilities and enable them to play a more active role in modern society.
In the UK, the Royal National Institute of Blind People (RNIB) has prioritised enabling people with visual impairments to use some of the services and products many people take for granted, such as public transport and cellphones~\cite{rnib-objectives}.
Improvements in modern computing have made it possible for new and innovative solutions for these problems to come to the fore.
In particular, researchers in the field of active vision have have made much progress in allowing a machine to autonomously manipulate a camera to gain more information about an environment to complete mapping and object and face tracking tasks, for example~\cite{bajcsy2018revisiting}.
There is, however, a significant research question about whether techniques from active machine vision can be applied to humans, i.e.\ can a machine identify a point of interest in a scene and direct a human, instead of an electronic servo, to focus a camera on that point? If this can be done, it would be beneficial to people with visual impairments and will allow them to search for an arbitrary point of interest or identify an unknown scene.

To this end, we are developing a mobile device-based navigation system using a Google Project Tango device\footnote{https://developers.google.com/tango}, pictured in \cref{fig:tango-headphone}, that caters to the perception limitations of people with limited vision by using a non-visual interface.
This system will ultimately allow users with visual impairments to reach a target location in the final leg of their journey, i.e.\ the so-called `last 10 yards' problem~\cite{google2016blind,bellotto2013}. 
The ideas here can also be extended and applied to existing macronavigation tools, such as Google Maps.
A Tango-enabled device comes pre-equipped with powerful image-processing, localisation and depth-perception capabilities and is built on top of a standard Android platform, providing access to the entire set of input/output options that this operating system has to offer.
The final system will use multiple feedback modes (vibration, audio and voice) to guide a user toward a target destination while providing information on any oncoming obstacles.

%\begin{figure}
  %\centering
  %\begin{subfigure}[t]{0.45\textwidth}
\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{figures/tango_headphone.png}
  \caption{The Tango device and bone-conducting headset used by the participants. }
  \label{fig:tango-headphone}
\end{figure}
  %\end{subfigure}\hfill
  %\begin{subfigure}[t]{0.45\textwidth}
\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{figures/camera_coordinate.pdf}
  \caption{A diagram showing the Tango's frame of reference and the target's pan ($\phi$) and tilt ($\theta$) angles.}
  \label{fig:cam-coords}
\end{figure}
  %\end{subfigure}
  %\caption{The hardware used during the experiment, as well as a diagram showing the Tango device's coordinate system. }
%\end{figure}

In this paper we focus on the audio feedback mode and discuss how it is used in our system.
We present experiments we performed to determine its effectiveness at directing a user to complete a pointing task and how its parameter values affect a user's target-search performance.
For the pointing task, we presented blindfolded and partially sighted participants with a set of virtual targets, one after another, and asked them to simply point a camera to where they thought the targets were.
The targets' pan and tilt angles w.r.t.\ the Tango's vertical plane, shown in \cref{fig:cam-coords}, were given to the participants through a spatial tone with varying pitch played via a set of bone-conducting headphones.
We use the latter to not interfere with the user's normal hearing function, which people with visual impairments typically rely on.
Because these headphones bypass the external structure of the ear responsible for localising a sound source's elevation, it is necessary to convey the target's tilt angle using another method.
Therefore, we opted for a tone with varying pitch to convey the angle adjustment required to point the camera in a particular direction.

Unfortunately, there is a gap in the literature regarding the use of a such an audio interface for pointing tasks with regards to a mobile device.
It is also unclear whether popular metrics, such as Fitts's Law~\cite{fitts1954information}, can be applied in this case.
Fitts's Law is a predictive model in the field of psychomotor research and often used in the human-machine interfacing field that relates the time it takes a user to direct a pointing device toward a target as a function of the difficulty of finding the target (i.e.\ the ratio between the distance to the target and its size).
If a Fitts relationship is found, it can be used as a metric to anticipate whether future adjustments to the pitch's rate of change will have a positive effect on target-search performance.

@@

The remainder of this paper is organised as follows: Section~\ref{sec:lit-review} gives an overview of existing navigation systems and audio interfaces for people living with visual impairments. Section~\ref{sec:portable-navigation} discusses the navigation device's hardware and interface setup and describes how the pan and tilt angles are conveyed to the user. The experiments that were conducted are explained in Section~\ref{sec:experiments} and the results are presented and discussed in Section~\ref{sec:results}. Finally, Section~\ref{sec:conclusion} concludes this paper with a summary of the findings and comments on our future work.

\section{Previous Work}
\label{sec:lit-review}

Attempting to deliver a system that allows people with visual impairments to independently navigate and accomplish everyday tasks is not new; in fact, there are multiple commercial systems and research prototypes already available on the market.
One approach that has been investigated is to outfit the existing white walking cane with various sensors,
such as sonar, radar, motor encoders, etc.~\cite{ulrich1997, marion2008batcane} to warn the user of upcoming obstacles, instead of relying on haptic feedback from the impact between the cane and the obstacle.
Another approach is to outfit a walking cane to act as a radio-frequency identification (RFID) antenna that can read a set of RFID tags placed at strategic locations around the environment at key spots or along a path~\cite{faria2010electronic, willis2005}.
This modification to the traditional cane is more discreet than the systems mentioned earlier and has been shown to work well.
Bluetooth tags and a smartphone can also be used in a similar manner~\cite{sato2017navcog3}.
However, the major drawbacks here are the significant cost and effort of modifying existing infrastructure with RFID and Bluetooth tags and maintaining them to stay up to date with a changing environment.
GPS-based systems~\cite{ran2004drishti, loomis2001navigating, kammoun2012navigation}, while cheap and reliable in outdoor environments, are not applicable in built-up urban areas and indoors where GPS signals are notoriously unreliable.
Computer vision-based systems provide a good compromise between usability, cost and accuracy and have been the focus of much research recently~\cite{manduchi2014last}.
One increasingly popular solution is to use RGB-D depth sensing cameras, which are becoming increasingly more accurate and cost-effective, to build a 3D image map of the environment, allowing a user to safely traverse through it~\cite{lee2015, rodriguez2012obstacle}.
Another approach is to use vision-based object recognition techniques to detect various objects and landmarks, such as doors, staircases, etc., and communicate their relative location to the user~\cite{schauerte2012assistive, tian2013b, fiannaca2014headlock}, or use audio instructions to instruct a user with vision impairment to fully capture an object of interest~\cite{vazquez2012helping}.

@@

%####################################################
%REVIEW SECTION - MAKE SURE TO MENTION DIFFERENCE BETWEEN BLIND AND SIGHTED , sonification of 3d scenes in an electronic travel aid for the blind.
Spatial audio has also been considered to convey the direction of a target.
Experiments have previously determined that people are able to find the location of a stationary sound source with an error of $\pm$\SI{35}{\degree}, in both the pan and tilt dimensions~\cite{zwiers2001spatial}, for both early-blind and normally-sighted people.
However, in~\cite{lewald2013exceptional, lessard1998early} it was found that the blind have a clear advantage in localisation accuracy over sighted people when presented with more difficult tasks, such as targets in motion and narrow-band stimuli, with~\cite{lewald2013exceptional} reporting an average absolute localisation error of around \SI{10}{\degree} in the pan and tilt dimensions.
Other authors found that the minimum difference in the spatial sound's angle for the user to be able to perceive movement is approximately \SI{1.7}{\degree}. During the experiments, an audio speaker was physically manoeuvred to provide the participant with a spatialised sound tone~\cite{ashmead1998spatial}.
Researchers have also tried using simulated spatial audio to inform the user from which direction a sound is originating~\cite{holland2002audiogps, kammoun2012navigation, rebillat2009smart, menelas2010audio, wilson2007swan, zotkin2004rendering}.
In these works, a sound was played through a set of headphones and the source spatialised with a head-related transfer function (HRTF), tricking the listener into thinking the sound source was located at some arbitrary 3D location.
Various audition techniques and methods were then used to guide the user along a path.
Other authors have also provided a framework for evaluating the quality of this 3D sound in terms of its psychoacoustic properties~\cite{guastavino2004perceptual, nicol2014roadmap}.
There are also experimental results about how well users can find targets presented with spatial sound in the tilt and panning dimensions with normal headphones~\cite{katz2011spatial, zwiers2001spatial}, as well as work that has shown similar levels of localisation performance between normal and bone-conducting headphones in the pan dimension~\cite{macdonald2006spatial}.
However, to our knowledge no extensive work or experiments have been done to determine how well users respond to tilt adjustment instructions using a tone with \emph{varying pitch}, in particular when using bone-conducting headphones. 
%########################################################

@@

\section{Portable Navigation System}
\label{sec:portable-navigation}

@@

The system we intend to ultimately deliver is a portable navigation device that is capable of guiding a user with visual impairments on the last leg of their journey, e.g.\ to a specific aisle and shelf in a shop.
To do this, we will use a combination of different feedback modes to facilitate two-way communication between the user and the device.
A large amount of data need to be translated from a visual form into a format that is useful to people with limited or no vision.
We therefore plan to use a combination of voice, audio and vibration cues to translate the visual navigation data as effectively as possible and overcome the bandwidth limitation of the human ear.

The system is based on a concept proposed in~\cite{bellotto2013, lock2017portable}, which uses a Google Tango device.
This is an Android-based cellphone or tablet device that comes equipped with an RGB-D camera to sense colour and depth.
It combines an inertial measurement unit with powerful and robust landmark recognition and image processing algorithms to localise itself.
An added benefit of this platform is its familiar, compact form-factor which will help overcome the hurdle of user-acceptance and usability.

We use a set of bone-conducting headphones (see \cref{fig:tango-headphone}) that are placed externally on the user's head so that the system does not occlude the perception of real-world sounds and does not interfere with a user's normal hearing function.
In the future, our system will use multiple feedback modes to provide the user with navigation and obstacle avoidance instructions.
However, for this paper, we only considered the spatialised audio mode and its variation in pitch in order to determine its effectiveness in conveying a target's pan and tilt angles.

@@

\begin{figure}
  \centering
  \includegraphics[clip=true, trim=50 120 80 50, width=0.4\textwidth]{figures/pipeline.pdf}
  \caption{A diagram of the individual system components and their communication pipelines. $F$ indicates a feedback signal and $P$ a pose signal. }
\label{fig:pipeline}
\end{figure}

@@

For the series of experiments performed in this work, only the audio feedback mode was used to interface with the user.
The audio component is responsible for conveying the 2D position of a target on the vertical plane w.r.t.\ the Tango's coordinate system, showed in \cref{fig:cam-coords}, in terms of pan and tilt angles.
Distance to the target can be rendered by adjusting the audio signal's amplitude as a function of the distance.
However, in this work, this was not implemented.
The audio signal is a sinusoidal sound wave that is continuously played (refreshed at $\sim$\SI{10}{\hertz}) to the user through bone-conducting headphones.
We select a sinusoidal wave because it is relatively simple to manipulate and analyse.
A richer and more natural tone will be implemented at a later stage of the project.

The audio is spatialised in the pan dimension using an HRTF provided by the OpenAL sound library\footnote{http://openal.org}, while the tilt angle is conveyed by varying the pitch of the audio tone.
We use this approach because the external set of bone-conducting headphones plays the sound through the user's cheekbones instead of their outer ears, bypassing the ears' pennae that provide humans their ability to localise elevated sound sources~\cite{roffler1968factors, algazi2001elevation}.
This makes an HRTF less effective in conveying the target's tilt angle and therefore needs to be conveyed using another method; a varying pitch tone in this case.
The difference between the target's angular position and the angular orientation of the Tango device are used to generate the audio signals.

@@

The pan angle describes the angle which the user needs to rotate the camera vector around the vertical $z$-axis in \cref{fig:cam-coords}, i.e.\ how far the target is to the left or right of the user.
We use an HRTF to add a spatial element to the audio tone that the system plays to the user, making the tone sound like it is coming from the direction of the target.
Humans determine the pan direction by subconsciously analysing the time difference between a sound reaching both of your ears.
For example, if you hear a sound in your left ear first, you interpret the sound as coming from your left.
The greater the so-called inter-aural time difference~\cite{wightman1992dominant}, the greater the perceived angular distance to the sound source.

We implement OpenAL's default HRTF, based on the MIT's KEMAR dataset\footnote{http://sound.media.mit.edu/resources/KEMAR.html}, to generate a sinusoidal sound wave based on the difference between the user and target's positions.
The library takes position values as input and outputs a tone based on the angle between the two position vectors.

@@

The system adjusts the tone's pitch to convey the target's tilt angle w.r.t.\ the camera's current pointing vector, $C$, as shown in \cref{fig:cam-coords}. 
Here, a high pitch means the target is above the camera vector (i.e.\ the user should look up) and a low pitch means the target is below the camera vector (the user should look down).
This high/low association scheme is chosen because humans naturally associate high-pitched sounds with higher objects and lower-pitched noises with lower objects~\cite{pratt1930spatial, blauert1997spatial}.
We use a logarithmic, octave-based gain function for the pitch, since an increase in octave provides a distinct perceptible change while keeping the timbre roughly constant~\cite{shepard1964circularity}.

We wish to determine how the gradient of the pitch gain function affects a user's performance.
For example, does an increased gain rate in the pitch as a function of the tilt angle lead to an increased target acquisition rate? For this we select three different pitch gain gradients, referred to here as low (\emph{lo}), medium (\emph{med}) and high (\emph{hi}) gain presets.
To find these gradients, we set the maximum and minimum limits for the tilt angle and the maximum and minimum frequencies for the pitch.
Furthermore, for the sake of consistency, each gradient setting is set to generate the same pitch when the user is pointing the camera towards the target.

After practical tests with the Tango and headphones, we set the neutral, on-target tone to a frequency of \SI{512}{\hertz} for its audibility.
For the \emph{lo} setting, we set the maximum and minimum pitches to be one octave higher and lower than the neutral tone, giving limits of \SI{1024}{\hertz} and \SI{256}{\hertz} respectively.
The \emph{med} preset is set to two octaves higher and lower than the neutral tone (\SI{2048}{\hertz} and \SI{128}{\hertz}) and the \emph{hi} to three octaves higher and lower (\SI{4096}{\hertz} and \SI{64}{\hertz}).
We selected these limits for practical reasons, given the fact that the bone conducting headphones we use have low volume gain at very high and low frequencies, making it difficult to hear.

The frequency is selected using Equation~\ref{eq:pitch-eq}, where $\theta$ is the tilt angle and $m$ and $c$ are parameters calculated using a linear equation (seen in Equation~\ref{eq:linear}) and the cut-off frequency values.
Here, $y_m$ is a cut-off frequency, $x_m$ its corresponding angle and $m$ the gradient for the setting, calculated as $m = \frac{\Delta y}{\Delta x}$.
The pitch gain functions are shown in \cref{fig:pitch-gain}.

\begin{equation}
  \label{eq:pitch-eq}
  f = 2^{-m \theta + c}
\end{equation}

\begin{equation}
  \label{eq:linear}
  y - y_m = m(x - x_m)
\end{equation}

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{figures/pitch_gain_functions.png}
  \caption{The pitch gain functions for the 3 different gain settings. }
  \label{fig:pitch-gain}
\end{figure}

The \SI{0}{\radian} position is directly in front of the user.
We set the range of angles the system can convey to $[-\frac{\pi}{2}; \frac{\pi}{2}]$; anything outside this range implies that the target is behind the user, and was therefore not considered in this series of experiments.

\section{Experiments}
\label{sec:experiments}

We performed a set of experiments with our audio interface to determine how effective it is at directing a user to adjust the pan and tilt angles of a camera for point it to a target.
Furthermore, we also carried out a set of pre-screening experiments to determine each participant's hearing characteristics.
The participants were given time before the experiments commenced to familiarise themselves with the device and the tones it emits, as well as what the `on-target' tone sounds like.
We also tested the system with 3 legally blind participants and compared their data to the blindfolded dataset.
Furthermore, these tests provided us with subjective feedback on the system that can be used to improve future iterations of the system.
The results from the experiments we performed allow us to better understand how the users respond to different settings for the spatial audio feedback stimulus and use them to improve and optimise the behaviour of the feedback modes in our portable navigation aid.

@@

For the experiments we used two groups of participants, with one including 42 sighted and blindfolded participants and the other 3 legally blind volunteers with varying degrees of vision impairment. 
They performed a series of experiments using our system with a pair of bone-conducting headphones.
The participants were recruited on a volunteer-basis and consisted of a diverse group of undergraduate students and professionals with ages ranging between 17 and 40 years (13 male, 32 female).
The participants also reported having no significant hearing issues or any other disability that could affect the experiment.
The participants participated in three experiments.
The first two experiments were performed to determine each participant's hearing characteristics and provide some context to the results generated during the target-search experiment.
These characterisation experiments will also indicate pre-existing biases our participant-base might have in the modes or dimensions we performed our third experiment on: the target search.

@@

The distance between the participant and the target is not considered here and the 3D targets are therefore generated at a constant distance of \SI{2}{\metre} from the participant.
Throughout the experiments, various parameters of the target and the participant are recorded and streamed in real-time to a laptop computer via a WiFi connection.
Prior to the target search experiment, the blindfolded participant group was given a few minutes without a blindfold to familiarise themselves with the system and confirm the target's location with their own eyes.
This was done verbally with the participants with limited vision.

\subsection{Participant Characterisation}

\subsubsection{Spatial Awareness}

>>>>

\begin{abstract}
	Our aim is to build a portable active navigation system for the visually impaired that uses active perception techniques and a combination of feedback modes to identify a scene and guide a user to their destination.
	Such a system requires a non-visual feedback interface and in this paper we investigate the effectiveness of a spatial audio tone with a varying pitch component, played through bone-conducting headphones, in conveying the pan and tilt angles of a target to the user in a pointing task.
	The goal is to determine how changes to the rate of change of the pitch affects a user's performance.
	For this, wee conducted a set of experiments with blindfolded and visually impaired users and found that the varying pitch component gives an average error of approximately \SI{20}{\degree} when searching for a target; in line with previous results from other authors.
	Furthermore, we are able to demonstrate that the audio interface adheres to Fitts's Law and used it as a metric to determine which pitch setting produces the optimal results.
	We discovered a trade-off between the speed and accuracy in the pointing task, which are maximised when the tone's pitch is adjusted to a low and high rate of change respectively. 
\end{abstract}

@@

In recent years, governments have spearheaded numerous initiatives to support people with disabilities and enable them to play a more active role in modern society.
In the UK, the Royal National Institute of Blind People (RNIB) has prioritised enabling people with visual impairments to use some of the services and products many people take for granted, such as public transport and cellphones~\cite{rnib-objectives}.
Improvements in modern computing have made it possible for new and innovative solutions for these problems to come to the fore.
In particular, researchers in the field of active vision have have made much progress in enabling machines to autonomously manipulate cameras to gain more information about an environment to complete mapping and object and face tracking tasks, for example~\cite{bajcsy2018revisiting}.
There is, however, a significant research question about whether techniques from active machine vision can be applied to humans, i.e.\ can a machine identify a point of interest in a scene and direct a human, instead of an electronic servo, to focus a camera on that point?
If this can be done, it would be beneficial to people with visual impairments and will augment their ability to search for an arbitrary point of interest or identify an unknown scene.

To this end, we are developing a mobile device-based navigation system using a Google Project Tango device\footnote{https://developers.google.com/tango}, pictured in \cref{fig:tango-headphone}, that caters to the perception limitations of people with limited vision by using a non-visual interface.
This system will ultimately allow users with visual impairments to reach a target location in the final leg of their journey, i.e.\ the so-called `last 10 yards' problem~\cite{google2016blind,bellotto2013}. 
The ideas here can also be extended and applied to mainstream navigation tools, such as Google Maps.
A Tango-enabled device comes pre-equipped with powerful image-processing, localisation and depth-perception capabilities and is built on top of a standard Android platform, providing access to the entire set of input/output options that this operating system has to offer.
The final system will use multiple feedback modes (vibration, audio and voice) to guide a user toward a target destination while providing information on any oncoming obstacles.

%\begin{figure}
  %\centering
  %\begin{subfigure}[t]{0.45\textwidth}
\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{figures/tango_headphone.png}
  \caption{The Tango device and bone-conducting headset used by the participants. }\label{fig:tango-headphone}
\end{figure}
  %\end{subfigure}\hfill
  %\begin{subfigure}[t]{0.45\textwidth}
  %\end{subfigure}
  %\caption{The hardware used during the experiment, as well as a diagram showing the Tango device's coordinate system. }
%\end{figure}

In this paper we focus on the audio feedback mode and discuss how it is used in our system.
We present experiments we performed to determine its effectiveness at directing a user to complete a pointing task and how its parameters affect users' target-search performance.
For the pointing task, we presented blindfolded and partially sighted participants with a set of virtual targets, one after another, and asked them to point a camera to where they thought the targets were.
The targets' pan and tilt angles ($\phi$ and $\theta$ respectively) w.r.t.\ the Tango's vertical plane, shown in \cref{fig:cam-coords}, were given to the participants through a spatial tone with varying pitch played via a set of bone-conducting headphones.
We use the latter to not interfere with the user's normal hearing function, which people with visual impairments typically rely on.
Because these headphones bypass the external structure of the ear responsible for localising a sound source's elevation, it is necessary to convey the target's tilt angle using another method.
Therefore, we opted for a tone with varying pitch to convey the angle adjustment required to point the camera in a particular direction.

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{figures/camera_coordinate.pdf}
  \caption{A diagram showing the Tango's frame of reference and the target's pan ($\phi$) and tilt ($\theta$) angles.}\label{fig:cam-coords}
\end{figure}

Unfortunately, there is a gap in the literature regarding the use of a such an audio interface for pointing tasks with regards to a mobile device.
It is also unclear whether popular metrics, such as Fitts's Law~\cite{fitts1954information}, can be applied in this case.
Fitts's Law is a predictive model in the field of psychomotor research, often used in the human-machine interfacing field, that relates the time it takes a user to direct a pointing device toward a target as a function of the difficulty of finding the target (i.e.\ the ratio between the distance to the target and its size).
If a Fitts relationship is found, it can be used as a metric to anticipate whether future adjustments to the pitch's rate of change will have a positive effect on target-search performance.

@@

The remainder of this paper is organised as follows: \cref{sec:lit-review} gives an overview of existing navigation systems and audio interfaces for people living with visual impairments.
\cref{sec:portable-navigation} discusses the navigation device's hardware and interface setup and describes how $\phi$ and $\theta$ are conveyed to the user.
The experiments that were conducted are explained in \cref{sec:experiments} and the results are presented and discussed in \cref{sec:results}. 
Finally, \cref{sec:conclusion} concludes this paper with a summary of the findings and comments on our future work.

\section{Previous Work}\label{sec:lit-review}

\todo{chop up this section. Make anonymous}
Attempting to deliver a system that allows people with visual impairments to independently navigate and accomplish everyday tasks is not new; in fact, there are multiple commercial systems and research prototypes already available on the market.
One approach that has been investigated is to outfit the existing white walking cane with various sensors,
such as sonar, radar, motor encoders, etc.~\cite{ulrich1997, marion2008batcane} to warn the user of upcoming obstacles, instead of relying on haptic feedback from the impact between the cane and the obstacle.
Another approach is to outfit a walking cane to act as a radio-frequency identification (RFID) antenna that can read a set of RFID tags placed at strategic locations around the environment at key spots or along a path~\cite{faria2010electronic, willis2005}.
This modification to the traditional cane is more discreet than the systems mentioned earlier and has been shown to work well.
Bluetooth tags and a smartphone can also be used in a similar manner~\cite{sato2017navcog3}.
However, the major drawbacks here are the significant cost and effort of modifying existing infrastructure with RFID and Bluetooth tags and maintaining them to stay up to date with a changing environment.
GPS-based systems~\cite{ran2004drishti, loomis2001navigating, kammoun2012navigation}, while cheap and reliable in outdoor environments, are not applicable in built-up urban areas and indoors where GPS signals are notoriously unreliable.
Computer vision-based systems provide a good compromise between usability, cost and accuracy and have been the focus of much research recently~\cite{manduchi2014last}.
One increasingly popular solution is to use RGB-D depth sensing cameras, which are becoming increasingly more accurate and cost-effective, to build a 3D image map of the environment, allowing a user to safely traverse through it~\cite{lee2015, rodriguez2012obstacle}.
Another approach is to use vision-based object recognition techniques to detect various objects and landmarks, such as doors, staircases, etc., and communicate their relative location to the user~\cite{schauerte2012assistive, tian2013b, fiannaca2014headlock}, or use audio instructions to instruct a user with vision impairment to fully capture an object of interest~\cite{vazquez2012helping}.

@@

Spatial audio has also been considered to convey the direction of a target.
Experiments have previously determined that people are able to find the location of a stationary sound source with an error of $\pm$\SI{35}{\degree}, in both the pan and tilt dimensions~\cite{zwiers2001spatial}, for both early-blind and normally-sighted people.
However, in~\cite{lewald2013exceptional, lessard1998early} it was found that the blind have a clear advantage in localisation accuracy over sighted people when presented with more difficult tasks, such as targets in motion and narrow-band stimuli, with~\cite{lewald2013exceptional} reporting an average absolute localisation error of around \SI{10}{\degree} in the pan and tilt dimensions.
Other authors found that the minimum difference in the spatial sound's angle for the user to be able to perceive movement is approximately \SI{1.7}{\degree}. During the experiments, an audio speaker was physically manoeuvred to provide the participant with a spatialised sound tone~\cite{ashmead1998spatial}.
Researchers have also tried using simulated spatial audio to inform the user from which direction a sound is originating~\cite{holland2002audiogps, kammoun2012navigation, rebillat2009smart, menelas2010audio, wilson2007swan, zotkin2004rendering}.
In these works, a sound was played through a set of headphones and the source spatialised with a head-related transfer function (HRTF), tricking the listener into thinking the sound source was located at some arbitrary 3D location.
Various audition techniques and methods were then used to guide the user along a path.
Other authors have also provided a framework for evaluating the quality of this 3D sound in terms of its psychoacoustic properties~\cite{guastavino2004perceptual, nicol2014roadmap}.
There are also experimental results about how well users can find targets presented with spatial sound in the tilt and panning dimensions with normal headphones~\cite{katz2011spatial, zwiers2001spatial}, as well as work that has shown similar levels of localisation performance between normal and bone-conducting headphones in the pan dimension~\cite{macdonald2006spatial}.
However, to our knowledge no extensive work or experiments have been done to determine how well users respond to tilt adjustment instructions using a tone with \emph{varying pitch}, in particular when using bone-conducting headphones.

@@

\section{Portable Navigation System}\label{sec:portable-navigation}

@@

We intend to ultimately deliver a portable navigation device that is capable of guiding a user with visual impairments on the last leg of their journey, e.g.\ to a specific aisle and shelf in a shop.
To do this, we will use a combination of different feedback modes to facilitate two-way communication between the user and the device.
A large amount of data need to be translated from a visual form into a format that is useful to people with limited or no vision.
We therefore plan to use a combination of voice, audio and vibration cues to translate visual navigation data as effectively as possible and overcome the bandwidth limitation of the human ear.

%The system is based on a concept proposed in~\cite{bellotto2013, lock2017portable}, which uses a Google Tango device.
This is an Android-based cellphone or tablet device that comes equipped with an RGB-D camera to sense colour and depth.
It combines an inertial measurement unit with powerful and robust landmark recognition and image processing algorithms to localise itself in real-time.
An added benefit of this platform is its familiar, compact form-factor which will help overcome the hurdle of user-acceptance and usability.

We use a set of bone-conducting headphones (see \cref{fig:tango-headphone}) that are placed externally on a user's head so that the system does not occlude the perception of real-world sounds and does not interfere with a user's normal hearing function.
In the future, we will use multiple feedback modes to provide the user with navigation and obstacle avoidance instructions.
However, for this paper, we only considered the spatialised audio mode and its variation in pitch in order to determine its effectiveness in conveying a target's pan and tilt angles.

@@

\begin{figure}
  \centering
  \includegraphics[clip=true, trim=50 120 80 50, width=0.4\textwidth]{figures/pipeline.pdf}
  \caption{A diagram of the individual system components and their communication pipelines. $F$ indicates a feedback signal and $P$ a pose signal. }\label{fig:pipeline}
\end{figure}

@@

For the series of experiments performed in this work, only the audio feedback mode was used to interface with the user.
The audio component is responsible for conveying the 2D position of a target on the vertical plane w.r.t.\ the Tango's coordinate system, showed in \cref{fig:cam-coords}, in terms of pan and tilt angles.
In future system iterations, the distance to a target can be rendered by adjusting the audio signal's amplitude as a function of the distance.
The audio signal is a continuous $\sim$\SI{10}{\hertz} sinusoidal sound wave that is played to the user through bone-conducting headphones.
We select a sinusoidal wave because it is relatively simple to manipulate and analyse.
A richer and more natural tone will be implemented at a later stage of the project.

The audio is spatialised in the pan dimension using the OpenAL\footnote{http://openal.org} sound library's HRTF, while the tilt angle is conveyed by varying the audio tone's pitch.
We use this approach because the external set of bone-conducting headphones plays the sound through a user's cheekbones instead of their outer ears, bypassing the ears' pennae that provide humans their ability to localise elevated sound sources~\cite{roffler1968factors, algazi2001elevation}.
This makes an HRTF less effective in conveying the target's tilt angle and therefore needs to be conveyed using another method; a varying pitch tone in this case.
The difference between the target's angular position and the device's angular orientation are used to generate the audio signals.

@@

The pan angle ($\phi$) describes the angle which the user needs to rotate the camera vector around the $z$-axis in \cref{fig:cam-coords}, i.e.\ how far the target is to the left or right of the user.
We use an HRTF to add a spatial element to the audio tone that is played to the user, making the tone sound like it is coming from the direction of the target.
Humans determine a sound source's pan direction by subconsciously analysing the time difference between a sound reaching both of your ears.
For example, if you hear a sound in your left ear first, you interpret the sound as coming from your left.
The greater the so-called inter-aural time difference, the greater the perceived angular distance to the sound source~\cite{wightman1992dominant}.

We implement OpenAL's default HRTF, based on the MIT's KEMAR dataset\footnote{sound.media.mit.edu/resources/KEMAR.html}, to generate a sinusoidal sound wave based on the difference between the user and target's positions.
The library takes position values as input and outputs a tone based on the angle between the two position vectors.

@@

The system adjusts the tone's pitch ($f$) to convey the target's tilt angle ($\theta$) w.r.t.\ the camera's current pointing vector, $C$, as shown in \cref{fig:cam-coords}. 
Here, a high pitch means the target is above the camera vector (i.e.\ the user should look up) and a low pitch means the target is below the camera vector (the user should look down).
This high/low association scheme is chosen because humans naturally associate high-pitched sounds with higher objects and lower-pitched noises with lower objects~\cite{pratt1930spatial, blauert1997spatial}.
We use a logarithmic, octave-based gain function for the pitch, since an increase in octave provides a distinct perceptible change while keeping the timbre roughly constant~\cite{shepard1964circularity}.

We wish to determine how the gradient of the pitch gain as a function of $\theta$ (i.e.\ $m = \frac{df}{d\theta}$) affects a user's performance.
For example, does steeper $\frac{df}{d\theta}$ lead to an increased target acquisition rate?
For this we select three different pitch settings, referred to henceforth as $m_l$, $m_m$ and $m_h$ for the smallest, middle and highest gain gradients respectively. 
To find these gradients, we set the maximum and minimum limits for $\theta$ and the maximum and minimum frequencies for $f$.
Furthermore, for the sake of consistency, each gradient setting is set to generate the same pitch when the user is pointing the camera towards the target.

After practical tests with the Tango and headphones, we set the neutral, on-target tone to $f_{512} =$ \SI{512}{\hertz} for its audibility.
For $m_l$, we set $f_{max}$ and $f_{min}$ to one octave higher and lower than the neutral tone, giving limits of \SI{1024}{\hertz} and \SI{256}{\hertz} respectively.
For $m_m$, the $f_{max}$ and $f_{min}$ are set to two octaves higher and lower than the neutral tone (\SI{2048}{\hertz} and \SI{128}{\hertz}) and $m_h$ to three octaves higher and lower (\SI{4096}{\hertz} and \SI{64}{\hertz}).
We selected these limits for practical reasons, given the fact that the bone conducting headphones we use have low volume gain at very high and low frequencies, making it difficult to hear. 
The values of $m_l, m_m, m_h$ are visualised in \cref{fig:pitch-gain}.
$f$ is given by \cref{eq:pitch-eq}, where $c$ can be found using the current pitch setting's pitch and angle limits.

\begin{equation}\label{eq:pitch-eq}
  f = 2^{-m \theta + c}
\end{equation}

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{figures/pitch_gain_functions.png}
  \caption{The pitch gain functions for the 3 different gain settings. }\label{fig:pitch-gain}
\end{figure}

The \SI{0}{\radian} position is directly in front of the user.
We set the range of angles the system can convey to $[-\frac{\pi}{2}; \frac{\pi}{2}]$; anything outside this range implies that the target is behind the user and was not considered in this series of experiments.

\section{Experiments}\label{sec:experiments}

We performed a set of experiments with our audio interface to determine its effectiveness at directing a user to adjust a camera's $\phi$ and $\theta$ to point it to a target.
%Furthermore, we also carried out a set of pre-screening experiments to determine each participant's hearing characteristics.
The results from the experiments allow us to better understand how the users respond to different settings for the spatial audio feedback stimulus and use them to improve and optimise the behaviour of the feedback modes in our portable navigation aid.

The participants were given time before the experiments commenced to familiarise themselves with the device and the tones it emits, as well as what the $f_{512}$ `on-target' tone sounds like.
We also tested the system with 3 legally blind participants and compared their data to the blindfolded dataset.
These tests in particular provided us with subjective feedback on the system that can be used to improve future iterations.

@@

For the experiments we used two groups of participants, with one including 42 sighted and blindfolded participants and the other 3 legally blind volunteers with varying degrees of vision impairment. 
They performed a series of experiments using our system with a pair of bone-conducting headphones.
The participants were recruited on a volunteer-basis and consisted of a diverse group of undergraduate students and professionals with ages ranging between 17 and 40 years (13 male, 32 female) and did not not have significant hearing issues or any other disability that could affect the experiment.

%The participants participated in three experiments.
%The first two experiments were performed to determine each participant's hearing characteristics and provide some context to the results generated during the target-search experiment.
%These characterisation experiments will also indicate pre-existing biases our participant-base might have in the modes or dimensions we performed our third experiment on: the target search.

@@

We are only evaluating the interface in the pan and tilt dimensions and the 3D targets are therefore generated at a constant distance of \SI{2}{\metre} from the participant.
Throughout the experiments, various parameters of the target and the participant are recorded and streamed in real-time to a laptop computer via a WiFi connection.
Prior to the target search experiment, the blindfolded participant group was given a few minutes without a blindfold to familiarise themselves with the system and confirm the target's location with their own eyes.
This was done verbally with the participants with limited vision.

%\subsection{Participant Characterisation}

%\subsubsection{Spatial Awareness}
